Serum Biomarkers of Nutrition Pre and Post CFTR Modulator Use in Children with Cystic Fibrosis

### 2022-04-15 Edith Zemanick/CF Nutrition/cf_nutrition.Rmd
flow chart working but needs editing

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(readxl)
library(performance)
library(mcp)
library(lmerTest)
library(broom.mixed)
library(plotly)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
if(Sys.info()["sysname"] == "Windows"){
  home_dir = "S:/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
}
knitr::opts_knit$set(root.dir = home_dir)
rm(home_dir)
source("~/GitHub/CF/get_genotype_class_severity.R")
```

```{r data cleaning}
# Import
df = read.csv("./Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_dataset.csv",na.strings = "")
# Only those on effective modulators
on_effective = unique(df$Patient.ID[which(df[,'Modulator'] %in% c('Kalydeco','Trikafta'))])
df = df %>% filter(Patient.ID %in% on_effective)
# Non-highy efective to "None
df$Modulator[!df$Modulator %in% c('Kalydeco','Trikafta')] = "None"
# Before or after
df$Before_After = df$Age.at.Test..Years. < df$Age.at.Start..years.
df$Before_After = factor(df$Before_After,levels = c(FALSE,TRUE),labels = c("After","Before"))
# Count number of measures 
num_values = df %>% group_by(Patient.ID,Before_After) %>% 
  summarise(n = sum(!is.na(X25OH.Vitamin.D)),.groups = "drop") %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = n)
num_values$`NA` = NULL
colnames(num_values)[2:3] = c("vit_measures_after","vit_measures_before")
df = left_join(df,num_values,by = "Patient.ID")
# At least two annual nutritional measurements prior to first modulator start date
two_measures = num_values$Patient.ID[which(num_values$`vit_measures_before` >= 2)]
# At least one measurement of nutritional markers and one weight/height measurement >= three months post-modulator
time_after = df %>% 
  filter(Age.at.Test..Years. >= (Age.at.Start..years. + 0.25),!is.na(X25OH.Vitamin.D)) %>%
  pull(Patient.ID) %>% unique(.)
# Filter out those without above criteria, get first modulator
df = df %>% filter(Patient.ID %in% intersect(two_measures,time_after)) %>%
  group_by(Patient.ID) %>% 
  mutate(first_mod = head(Modulator[Modulator != "None"],1)) %>% ungroup()
# Genotype group
geno = read_excel("./CF Info/Sweat Tests from Internal Database_genotype class.xlsx")
geno = geno[,c('PatientID','genoRisk')]
df = left_join(df,geno,by = c('Patient.ID' = 'PatientID'))
# Time prior to first modulator
years = df %>% group_by(Patient.ID,Before_After) %>% 
  mutate(years_to_start_max = max(abs(Age.at.Test..Years. - Age.at.Start..years.))) %>% 
  select(Patient.ID,Before_After,years_to_start_max) %>% 
  ungroup() %>% distinct(.) %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = years_to_start_max)
years$`NA` = NULL
colnames(years)[2:3] = c('max_years_to_start','max_years_from_start')
df = left_join(df,years,by = "Patient.ID")
df$years = df$Age.at.Test..Years. - df$Age.at.Start..years.
```

@@

# Vitamin measures before and after modulator initiation

For each participant, we selected the pre and post visit closest to modulator start (and at least 3 months after start) with both a vitamin A and FEV1 % predicted value measure. Some visits were on different days and had different measures, so some missing values were filled in with the closest value (within 1 month).

```{r results='asis'}
nutrition_vars = c("VitaminA_Retinol","X25OH.Vitamin.D","Vitamin.E.Alpha",
                   "BMI","FEV1...pred")
# Get the most recent 
paired_df = df %>% 
  select(Patient.ID,Before_After,years,first_mod,all_of(nutrition_vars)) %>% 
  filter(years >= 0.25 | years < 0) %>%
  mutate(years = round(years,1)) %>%
  group_by(Patient.ID,years) %>%
  fill(all_of(nutrition_vars),.direction = "downup") %>%
  drop_na(any_of(c("VitaminA_Retinol"))) %>%
  group_by(Patient.ID,Before_After) %>%
  slice_min(abs(years)) %>% distinct() %>% 
  arrange(Patient.ID,abs(years)) %>%filter(row_number()==1) 
paired_df$Before_After = relevel(paired_df$Before_After,ref = "Before")

f = as.formula(paste0("Before_After ~ ",paste0(nutrition_vars,collapse = "+")))
p = paired(f,data = paired_df,id = Patient.ID)
summary(p,
        labelTranslations = list(VitaminA_Retinol = "Vitamin A Retinol",
                                 X25OH.Vitamin.D = "25OH-Vitamin D",
                                 Vitamin.E.Alpha = "Vitamin E Alpha",
                                 first_mod = "First Modulator"))
```

@@

```{r}
# Set up data again
vit = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report Vitamins_throughDecember2021.xlsx")
vit = vit %>% select(`Patient ID`,Date,VitaminA_Retinol:`Vitamin E Gamma`)
# New age at start
mods = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report CFTR Modulators_throughDecember2021.xlsx")
mods = mods[mods$Modulator %in% c('Kalydeco','Trikafta'),]
mods = mods %>% select(`Patient ID`,Modulator,`Earliest Date in Registry`) %>% 
  pivot_wider(id_cols = `Patient ID`,names_from = Modulator,
              values_from = `Earliest Date in Registry`)
colnames(mods)[2:ncol(mods)] = paste0(colnames(mods)[2:ncol(mods)]," Date")
mods$`First Mod Date` = pmin(mods$`Trikafta Date`,mods$`Kalydeco Date`,na.rm = T)
mods$`Second Mod Date` = pmax(mods$`Trikafta Date`,mods$`Kalydeco Date`)
# Add pancreatic status
pancreatic = read_excel("./CF Info/Sweat Tests from Internal Database_genotype class_withPancStatus.xlsx")
pancreatic = pancreatic %>% rename(`Patient ID` = PatientID) %>% select(`Patient ID`,PancreaticStatus)
# FEV1 and BMI
fev = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report_PFTs_ThroughDecember2021.xlsx")
fev = fev %>% select(PatientID,Date,BMI,`FEV1 % pred`) %>% rename(`Patient ID` = PatientID)
# Merge
df = full_join(vit,fev) %>% filter(`Patient ID` %in% mods$`Patient ID`,)
df = left_join(df,mods)
df = left_join(df,pancreatic)
# Time relative to first mod
df$time = as.numeric(difftime(df$Date,df$`First Mod Date`,units = "days"))/365.25
# Calculate spline things
df$c1 = 
  (df$time-0)*(df$time>=0)
df$c2 = 
  (as.numeric(difftime(df$Date,df$`Second Mod Date`,units = "days"))/365.25)*(df$Date >= df$`Second Mod Date`)
# Limit to 3 years prior, select columns
df = df %>% filter(!is.na(time) & time >= -3) %>% distinct()
# Vitamin A to numeric
df$VitaminA_Retinol = suppressWarnings(as.numeric(df$VitaminA_Retinol))
```

>>>>

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(readxl)
library(performance)
library(mcp)
library(lmerTest)
library(broom.mixed)
library(plotly)
library(knitr)
library(DiagrammeR)
library(lubridate)
knitr::opts_chunk$set(echo = FALSE)
if(Sys.info()["sysname"] == "Windows"){
  home_dir = "S:/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
}
knitr::opts_knit$set(root.dir = home_dir)
rm(home_dir)
source("~/GitHub/CF/get_genotype_class_severity.R")
```

# Data cleaning

```{r}
# Vitamins
vit = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report Vitamins_throughDecember2021.xlsx")
vit = vit %>% select(`Patient ID`,Date,VitaminA_Retinol:`Vitamin E Gamma`)
# New age at start
mods = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report CFTR Modulators_throughDecember2021.xlsx")
mods = mods[mods$Modulator %in% c('Kalydeco','Trikafta'),]
mods = mods %>% select(`Patient ID`,Modulator,`Earliest Date in Registry`) %>% 
  pivot_wider(id_cols = `Patient ID`,names_from = Modulator,
              values_from = `Earliest Date in Registry`)
colnames(mods)[2:ncol(mods)] = paste0(colnames(mods)[2:ncol(mods)]," Date")
mods$`First Mod Date` = pmin(mods$`Trikafta Date`,mods$`Kalydeco Date`,na.rm = T)
mods$`Second Mod Date` = pmax(mods$`Trikafta Date`,mods$`Kalydeco Date`)
# Add pancreatic status
pancreatic = read_excel("./CF Info/Sweat Tests from Internal Database_genotype class_withPancStatus.xlsx")
pancreatic = pancreatic %>% rename(`Patient ID` = PatientID) %>% select(`Patient ID`,PancreaticStatus)
# FEV1 and BMI
fev = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report_PFTs_ThroughDecember2021.xlsx")
fev = fev %>% select(PatientID,Date,BMI,`FEV1 % pred`) %>% rename(`Patient ID` = PatientID)
# Merge
df = full_join(vit,fev,by = c("Patient ID", "Date")) %>% filter(`Patient ID` %in% mods$`Patient ID`)
df = left_join(df,mods,by = "Patient ID")
df = left_join(df,pancreatic,by = "Patient ID")
```

```{r}
# Counts from Elin
# patients with CF in our clinical database from 2008- December 2021: 740 total, 474 in our dataset
ncdb = 740
# of patients treated with ivacaftor from 2008-2021: 58 total, 47 in dataset
# of patients treated with trikafta (ETI) from 2008-2021: 228 total, 150 in dataset 
# of patients treated with ivacaftor followed by ETI – these patients may be counted in both groups: 22 total, 17 in dataset
# Tim counts
df$Group = apply(df,1,function(r){
  if(!is.na(r["Kalydeco Date"]) & is.na(r["Trikafta Date"])){
    g = "Kalydeco"
  } else if (is.na(r["Kalydeco Date"]) & !is.na(r["Trikafta Date"])){
    g = "Trikafta"
  } else if (!is.na(r["Kalydeco Date"]) & !is.na(r["Trikafta Date"])){
    g = "Both"
  }
  return(g)
})
# Two nutrition measures before first mod start
two_measures = df %>% 
  mutate(before_first = Date < `First Mod Date` & !is.na(VitaminA_Retinol)) %>% 
  group_by(`Patient ID`,Group) %>% count(before_first) %>% 
  filter(before_first == T,n >= 2)
# At least one nutritional measure and one weight/height measure >= 3 months after first
after_measure = df %>% 
  mutate(after_first = (Date >= `First Mod Date` %m+% months(3)) & 
           !is.na(BMI) & 
           !is.na(VitaminA_Retinol)) %>%
  group_by(`Patient ID`,Group) %>% count(after_first) %>%
  filter(after_first == T,n >= 1)
# Filter 
df = df %>% filter(`Patient ID` %in% two_measures & `Patient ID` %in% after_measure)
```

```{r}
# Flowchart
grViz("
digraph graph2 {

graph [layout = dot]

# node definitions with substituted label text
node [shape = rectangle, width = 4, fillcolor = Biege]
a [label = '@@1']
b [label = '@@2']
c [label = '@@3']
d [label = '@@4']
e [label = '@@5']
f [label = '@@6']
g [label = '@@7']
h [label = '@@8']

a -> b
b -> c
c -> d
c -> e
c -> f
}

[1]: paste0('Clinical Database (n = ', ncdb, ')')
[2]: paste0('Treated with ivacaftor or ETI (or both) from 2008-2021 (n = ', length(unique(mods$`Patient ID`)),')')
[3]: paste0('In our dataset (n = ',length(unique(df$`Patient ID`)),')')
[4]: paste0('Treated with ivacaftor alone from 2008-2021 (n = ', n_kaly,')')
[5]: paste0('Treated with ETI alone from 2008-2021 (n = ', n_tri,')')
[6]: paste0('Treated with ivacaftor AND ETI from 2008-2021 (n = ',')')
[7]: paste0('At least two Vitamin A measures (n = ',')')
[8]: paste0('At least two Vitamin A measures (n = ',')')
[9]: paste0('At least two Vitamin A measures (n = ',')')
")
```

@@

@@

```{r}
# Time relative to first mod
df$time = as.numeric(difftime(df$Date,df$`First Mod Date`,units = "days"))/365.25
# Calculate spline things
df$c1 = 
  (df$time-0)*(df$time>=0)
df$c2 = 
  (as.numeric(difftime(df$Date,df$`Second Mod Date`,units = "days"))/365.25)*(df$Date >= df$`Second Mod Date`)
# Limit to 3 years prior, select columns
df = df %>% filter(!is.na(time) & time >= -3) %>% distinct()
```
